---
title: 'Chapter 3: Sampling the Imaginary'
author: "Quang Nguyen"
date: "8/8/2020"
output:
  html_document: default
  pdf_document: default
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
library(rethinking)
library(glue)
```

# Chapter 3: Sampling the Imaginary  

Easy problems rely on the following code: 
```{r easy_setup}
p_grid <- seq(from = 0, to = 1, length.out = 1000)
prior <- rep(1,1000)
likelihood <- dbinom(6, size = 9, prob = p_grid)
posterior <- likelihood * prior
posterior <- posterior / sum(posterior)
set.seed(100)
samples <- sample(p_grid, prob = posterior, size =1e4, replace = T)
```

### 3E1. 
Posterior probability that lies below $p = 0.2$
```{r 3e1}
sum(posterior[p_grid <0.2])
```
### 3E2
Posterior probability that lies above $p = 0.8$
```{r 3e2}
sum(posterior[p_grid > 0.8])

```

### 3E3  
Posterior probability lies between $p = 0.2$ and $p = 0.8$  
```{r 3e3}
sum(samples < 0.8 & samples > 0.2)/1e4
```
### 3E4  
20% of the posterior probability lies below $p$ value of 
```{r 3e4}
quantile(samples, 0.2)
```
### 3E5  
20% of the posterior probability lies above the $p$ value of 
```{r 3e5}
quantile(samples, 0.8)
```
### 3E6  
Values of $p$ containing the narrowest interval equal to 66% of the posterior probability  
```{r 3e6}
rethinking::HPDI(samples = samples, prob = 0.66)
```

### 3E7  
Values of $p$ containing the interval with 66% of the posterior probability assuming equal posterior probability both below and above the interval    
```{r 3e7}
rethinking::PI(samples = samples, prob = 0.66)
```
### 3M1  
Posterior distribution of globe tossing problem using grid approximation using new data of 8 water in 15 tosses
```{r 3m1}
p_grid <- seq(0,1,length.out = 1000)
prior <- rep(1,1000)
likelihood <- dbinom(x = 8, size = 15, prob = p_grid)
posterior <- prior * likelihood 
posterior <- posterior / sum(posterior)
plot(density(posterior))
```

### 3M2  
Using 10,000 samples to calculate the 90% HPDI for $p$  
```{r 3m2}  
set.seed(100)
samples <- sample(p_grid, size = 1e4, replace = T, prob = posterior)
rethinking::HPDI(samples, prob = 0.9)
hist(samples)
```
### 3M3  

